---
title: "Statistiques SD Project"
author: "Alice Clément & Nino Pujals"
date: "2025-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The objective of this report is to provide an analysis of the *Rossi* dataset which analyses the time before recidive of criminals in the USA depending on numerous covariates. 

The main question is the following: **Does a financial help reduce the risk of recidive ?**

# Preparing dataset

## Load packages

```{r warning=FALSE, message=FALSE, error=FALSE, results='hide'}
packages_names = c(
  'tidyverse',
  'survival',
  'ggsurvfit'
)

lapply(packages_names, library, character.only = TRUE)
```

## Load dataset

```{r}
Rossi <- read.csv("Rossi.csv", sep = ";", row.names=1)
```

## Inspect dataset

### General information
```{r examine data}
head(Rossi)

dim_Rossi = dim(Rossi)
cat(paste("The dataset contains", dim_Rossi[1], "samples and", dim_Rossi[2], "features."))

col_names = colnames(Rossi)
cat(paste(c("The features are:", col_names, ".")))

unique_dt = unique((lapply(Rossi, class)))
cat(paste("The features of the dataset have the following datatypes:", paste(unique_dt) , "."))
```

Here we can see that all the features of type `emp` correspond of being employed during the said week. The `NA` values correspond to weeks that occurred after the first arrest.

### New variable creation

Let us create a new variable to count the number of weeks of employment:

```{r creating new variable}
# extracting useful columns
col_names = unlist(paste("emp", as.character(1:52), sep=""))
col_names_ind = which(colnames(Rossi) %in% col_names)

# converting "yes"/"no" to 1/0
convert_yes_no <- function(x){
  ifelse(x == "yes", 1, 0)
}
Rossi <- Rossi %>% 
  mutate(across(all_of(col_names_ind), convert_yes_no)) 

# creating new feature
Rossi <- Rossi %>%
  mutate(nb_emp = rowSums(Rossi[col_names_ind], na.rm=TRUE))

# removing useless features
Rossi <- Rossi %>%
  select(-all_of(col_names_ind))

head(Rossi)
```

## Missing values

```{r missing values search}
nb_mv = sum(is.na(Rossi))
pr_mv = nb_mv / prod(dim_Rossi) * 100
cat(paste("The number of missing values inside the dataset is:", nb_mv, "which makes for", pr_mv,"% of the dataset."))
```

There is no missing values in the dataset, thus it is ready to be used.

# Survival model

## Financial aid

```{r}
# Adjust survival model with stratification
surv_model <- survfit(Surv(week, arrest) ~ fin, data = Rossi)

# plot with ggsurvfit
ggsurvfit(surv_model) +
  labs(
    x = "Weeks",
    y = "Overall survival probability",
    title = "Survival Curves by fin"
  ) +
  add_confidence_interval() +
  theme_minimal()
```

Here we can see that ex-prisoners that receive a financial aid upon their release have a higher survival function than those who do not.

## Race

```{r}
# Ajuster le modèle de survie avec stratification
surv_model <- survfit(Surv(week, arrest) ~ race, data = Rossi)

# Tracer les courbes avec ggsurvfit
ggsurvfit(surv_model) +
  labs(
    x = "Weeks",
    y = "Overall survival probability",
    title = "Survival Curves by fin"
  ) +
  add_confidence_interval() +
  theme_minimal()
```

Here we can see that ex-prisoners that are of black ethnicity have a lower survival function than those who are not.

## Accelerated Feature Time Model

```{r}
aft_model <- survreg(formula = Surv(week, arrest) ~ fin + age + race + wexp + mar + paro + prio + educ + nb_emp, data = Rossi)
summary(aft_model)
```

We consider only the features with significantly low p-value. 

Each week of employment during the year after the release increases the time before arrest by 3%. 

Each prior arrest decreases the time before arrest by 4%. 
